---
title: "DA5020.A9.Sheetal.Nighut"
author: "Sheetal Nighut"
date: "7/24/2022"
output: html_document
---

```{r}
# loading essential libraries

library(lubridate)
library(dplyr)
library(ggplot2)
library(scales)
```

# 1) Load the data into your R environment (found in Canvas). Ensure that you inspect the data, so that you know how to identify the necessary columns.


```{r}
# Read the csv file and importing it

air_pass <- read.csv("/Users/sheetalnighut/Desktop/DA_5020/Air_Traffic_Passenger_Statistics.csv")
head(air_pass)
```

```{r}
# Check the rows and columns 

dim(air_pass)

str(air_pass)
```



# 2) Filter the dataset to extract all domestic passenger activity that occurred each year, in the month of 
March. After which calculate the total passengers for each period. Visualize the extracted data, using a 
line chart; indicate the year on the x-axis and the total passengers on the y-axis. Comment on the 
visualization. Note: the final/aggregated dataset should have one row for March of each year. 


```{r}
# Filte the data set to get the domestic flights only
domestic_only <- air_pass[which(air_pass$GEO.Summary == "Domestic"),]

# Add the date to the Activity Period
domestic_only$Activity.Period <- paste0(domestic_only$Activity.Period, '01')

# Convert the Activity Period to year, month and day format
domestic_only$Activity.Period <- as.Date(domestic_only$Activity.Period, format = "%Y%m%d")

# Extract the columns having month = "March"
march_only <- domestic_only[which(months(domestic_only$Activity.Period) == 'March'),]

# In the domestic flights for the month of March, count the number of passengers each year
total_count <- march_only %>%
  mutate(Activity.Period =  year(march_only$Activity.Period))%>%
  group_by(Activity.Period) %>%
  summarise(passengers_cnt = sum(Passenger.Count))
head(total_count)
```


```{r}
# using ggplot visualize the data

ggplot(total_count, aes(x = Activity.Period, y = passengers_cnt, fill = Activity.Period)) +
  geom_line(stat = "Identity")+ xlab("Year") + ylab("Total domestic passengers") +
  labs(title = "All domestic passenger activity that occurred each year, in the month of March", caption = "Air_Traffic_Passenger_Statistics.csv")
```



# 3) Forecast the total passenger activity for March 2019, using a simple moving average of the following time 
periods: 201603, 201703 and 201803. After which, calculate the error (i.e. the difference between the 
actual and the predicted values for March 2019). Evaluate the results; how does it compare to the actual 
value for the total passenger count in March 2019? 

```{r}

# Extract the data for the periods 201603, 201703 and 201803. 
x <- total_count[11:13,]

# Move average to forecast the passenger activity for March 2019
mov_avg <- mean(x$passengers_cnt)
mov_avg
```
```{r}
# Actual value in 2019
total_count[14,]
```


# 4) Forecast the total passenger activity for 2019, using a three year weighted moving average. Apply the 
following weights: 3, 5, and 7 for the respective time periods: 201603, 201703 and 201803. After which, 
calculate the error and evaluate the result from your prediction. How does it compare to the actual value 
for the total passenger count in March 2019? 

```{r}
# Create a vector of weights
w <- c(3,5,7)

# Assign the weight to the count of passengers for the periods:  201603, 201703 and 201803. 
sw <- w* x$passengers_cnt

# Weighted Moving Average
sum(sw)/sum(w)
```
```{r}
# Actual value in 2019
total_count[14,]
```


# 5) Forecast the total passenger activity for 2019 using exponential smoothing (alpha is 0.7). Comment on the 
prediction for March 2019 with the actual value in the dataset. Note: use data from 2008 to 2018 to build 
your model. 

```{r}
#  data from 2008 to 2018 
lm_data <- total_count[3:13,]

# Set alpha as 0.7
alpha <- 0.7

# Create two columns Forecast and Error in the data frame
lm_data$Ft<- 0
lm_data$E <- 0

# Assign the value to first row of newly created columns
lm_data$Ft[1] <- as.numeric(lm_data[1,2])

# Assign the value to all rows of the columns

for (i in 2:nrow(lm_data)) {
  lm_data$Ft[i] <- as.numeric(lm_data$Ft[i-1] + alpha*lm_data$E[i-1])
  lm_data$E[i] <- as.numeric(lm_data[i,2] - lm_data$Ft[i])
}

#  Forecast the total passenger activity for 2019 using exponential smoothing 
forecast <- lm_data[3:13,]
n <- nrow(forecast)
F.es <- lm_data$Ft[n] + alpha*lm_data$E[n]
F.es

```

# 6) Build a simple linear regression model using the year and total passenger activity for all data from 2008 to 2018. After which, forecast the total passenger activity for 2019 and 2020. Comment on the results.

```{r}
# Add the date to the Activity Period
air_pass$Activity.Period <- paste0(air_pass$Activity.Period, '01')

# Convert the Activity Period to year, month and day format
air_pass$Activity.Period <- as.Date(air_pass$Activity.Period, format = "%Y%m%d")

# Group the data by year and the total count per year
extract<- air_pass%>% select(Activity.Period, Passenger.Count)%>%
  group_by(Years= year(Activity.Period))%>%
             summarise(passengers_cnt = sum(Passenger.Count))

# Filter the years from 2008 to 2018
q6 <- extract%>%filter(extract$Years %in% 2008:2018)
q6

# Build a simple linear regression model
model <- lm(passengers_cnt ~ Years, data= q6)
summary(model)
```

```{r}
# Forecast for the year 2019 
f_19 <- -66775833 - 35737*(2019)
f_19

# Forecast for the year 2019 
f_20 <- -66775833 - 35737*(2020)
f_20
```



# 7) Calculate the mean squared error (MSE) for the models in (5 and 6) above based on the data from 2008 to 
2018. Perform this step by step, using each model to make a forecast for each given time period, then 


```{r}
# Initialize columns for prediction and error
q6$P <- 0
q6$SE <- 0

# squared error value calculation
for (i in 1:nrow(q6)){
  q6$P[i] <- -8.656e+09 +4.346e+06*(q6$Years[i])
 q6$SE[i] <- (as.numeric(q6[i,2] -q6$P[i])^2)
}
head(q6)
```

```{r}
# Finding the MSE
format(mean(q6$SE), scientific = FALSE)
```

```{r}
# Intialize  the column of error to Exponential smoothing model data

 lm_data$SE <- 0

# Calculating squared error value

for(i in 1:nrow(lm_data)){
  lm_data$SE[i] <- (lm_data$E[i])^2
}

lm_data
```

```{r}
# Calculate MSE value for Exponential smoothing model 

format(mean(lm_data$SE), scientific= FALSE)
```

The MSE value for a linear regression model is 3543073376864 and MSE for Exponential smoothing model is 139638420983. MSE for is low  regression model. Hence, our linear regression model is more accurate as  it's MSE is lower than that of exponential smoothing model.


